---
title: PLAN 372 HW 6
author: Megan Gao
---

```{r}
library(tidyverse)
```

```{r}
trees <- read_csv("TS3_Raw_tree_data.csv")
```

## Question 1: Sample sizes by state

```{r}
trees <- trees %>%
  mutate(
    state = str_extract(City, "[A-Z]{2}$"),
    city = str_extract(City, "^[^,]+")
  )

state_counts <- trees %>%
  group_by(state) %>%
  summarize(n = n()) %>%
  ungroup()

state_counts
```

The dataset contains observations from multiple states, with California having the highest count, followed by other states across the US.

```{r}
ggplot(state_counts, aes(x = state, y = n)) +
  geom_col() +
  theme_minimal()
```

## Question 2: Cities in NC/SC

```{r}
trees_nc_sc <- trees %>%
  filter(state %in% c("NC", "SC"))

unique(trees_nc_sc$city)
```

## Question 3: Genera and species

```{r}
trees_nc_sc <- trees_nc_sc %>%
  mutate(genus = str_extract(ScientificName, "^[A-Z][a-z]+"))

genus_crown <- trees_nc_sc %>%
  group_by(genus) %>%
  summarize(mean_diameter = mean(`AvgCdia (m)`, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(mean_diameter))

genus_crown
```

The genus with the largest average crown diameter is `r genus_crown$genus[1]` with a diameter of `r round(genus_crown$mean_diameter[1], 2)` meters.

## Question 4: Tree age

```{r}
genus_age <- trees_nc_sc %>%
  group_by(genus) %>%
  summarize(mean_age = mean(Age, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(mean_age))

genus_age
```

Yes, there are differences in average age across genera. The oldest genus on average is `r genus_age$genus[1]` with a mean age of `r round(genus_age$mean_age[1], 1)` years. This could partially explain the crown diameter results, as older trees tend to have larger crowns.

## Extra Credit: Genera recommendation

```{r}
crown_per_year <- trees_nc_sc %>%
  filter(!is.na(Age) & Age > 0 & !is.na(`AvgCdia (m)`)) %>%
  group_by(genus) %>%
  summarize(
    mean_diameter = mean(`AvgCdia (m)`, na.rm = TRUE),
    mean_age = mean(Age, na.rm = TRUE),
    growth_rate = mean_diameter / mean_age,
    n = n()
  ) %>%
  ungroup() %>%
  filter(n >= 10) %>%
  arrange(desc(growth_rate))

crown_per_year
```

I recommend **`r crown_per_year$genus[1]`** as it has the highest crown diameter per year of age (`r round(crown_per_year$growth_rate[1], 3)` meters/year) among genera with at least 10 observations. This genus produces a large crown quickly.

## Extra Credit: Species

```{r}
trees <- trees %>%
  mutate(
    genus = str_extract(ScientificName, "^[A-Z][a-z]+"),
    species = str_extract(ScientificName, "(?:\\s+x\\s+|\\s+)[a-z]+") %>%
      str_remove("^\\s+x\\s+|^\\s+")
  )

species_per_genus <- trees %>%
  filter(!is.na(genus) & !is.na(species)) %>%
  group_by(genus) %>%
  summarize(n_species = n_distinct(species)) %>%
  ungroup() %>%
  arrange(desc(n_species))

species_per_genus
```

## AI Reflection

I used ChatGPT to help understand how to apply regular expressions and tidyverse functions such as mutate, filter, and summarize. I mainly used it to clarify syntax and debug issues. It supported my learning by helping me better understand how to manipulate strings and structure my analysis using concepts from class.